---
title: "Conflict, vaccination, and outbreaks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conflict, vaccination, and outbreaks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette demonstrates how to extend the core measles panel with conflict
exposure indicators and test whether conflict predicts:

1. next-year outbreak occurrence,
2. changes in vaccine coverage,
3. changes in measles case burden.

The full workflow runs in R and uses the package's bundled fixture data.

## 1) Build a panel with conflict indicators

```{r}
library(vpdsus)
library(dplyr)
library(ggplot2)

panel <- vpdsus_example_panel()

conf <- get_conflict(
  years = sort(unique(panel$year)),
  countries = unique(panel$iso3),
  source = "ucdp_fixture"
)

panel_conf <- merge_conflict_with_panel(panel, conf)

dplyr::summarise(
  panel_conf,
  n_rows = dplyr::n(),
  n_countries = dplyr::n_distinct(iso3),
  year_min = min(year, na.rm = TRUE),
  year_max = max(year, na.rm = TRUE),
  share_any_conflict = mean(conflict_any_event, na.rm = TRUE),
  mean_fatalities_per_100k = mean(conflict_fatalities_per_100k, na.rm = TRUE)
)
```

Conflict predictors used in this vignette:

- `conflict_any_event`: discrete indicator (0/1).
- `conflict_fatalities_per_100k`: continuous intensity indicator.

## 2) Prepare lagged modelling data

```{r}
conf_mod <- make_conflict_modelling_panel(
  panel_conf,
  lag_years = 1,
  outcome = "outbreak_pc",
  require_conflict_available = TRUE
)

dplyr::glimpse(conf_mod)
```

The modelling panel aligns year `t` predictors with year `t + 1` outcomes.

## 3) Fit conflict-effect models

```{r}
fit_conf <- fit_conflict_effect_models(conf_mod)

fit_conf$coefficients |>
  dplyr::filter(grepl("^conflict_", term)) |>
  dplyr::arrange(model, term)
```

Model outputs include:

- `model_outbreak`: logistic model for next-year outbreak occurrence.
- `model_delta_coverage`: linear model for next-year coverage change.
- `model_delta_cases`: linear model for next-year change in log cases.

## 4) Visualise conflict exposure over time

```{r}
conf_ts <- panel_conf |>
  dplyr::filter(conflict_data_available) |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    any_conflict_share = mean(conflict_any_event, na.rm = TRUE),
    mean_fatalities_per_100k = mean(conflict_fatalities_per_100k, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(conf_ts, aes(x = year, y = any_conflict_share)) +
  geom_line(linewidth = 0.8, colour = "#B22222") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Year",
    y = "Share of countries with conflict",
    title = "Global conflict occurrence in the panel"
  ) +
  theme_minimal(base_size = 11)
```

## Interpretation notes

- Treat this as a baseline explanatory analysis, not causal identification.
- Discrete and continuous conflict indicators can move differently; inspect both.
- Use rolling-origin validation from the outbreak vignette when making predictive claims.

