---
title: "Mechanistic modelling in odin2 (annual SIRV)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mechanistic modelling in odin2 (annual SIRV)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

It answers three practical questions:

1. What does each mechanistic helper do?
2. How does the discrete susceptible-balance model relate to Method D?
3. How can we run and interpret an annual SIRV model?

## 1) Mechanistic function map

- [odin2_build_model()]: load a compiled model generator.
- [odin2_simulate()]: run a model over time.
- [odin2_balance_inputs_from_panel()]: convert panel columns to model inputs.
- [calibrate_rho_case_balance()]: solve for reporting fraction `rho` to match a target susceptible count.

Model choices in this package:

- `balance_discrete`: one-state susceptible recurrence (teaching bridge to Method D).
- `sirv_annual`: annual `S`, `I`, `R`, `V` model with births, vaccination, infection, recovery, mortality, and reported cases output.

```{r}
library(vpdsus)
```

## 2) Minimal discrete model

Start with a one-state difference equation model:

- `S[t+1] = S[t] + inc`

```{r}
mdl <- odin2_build_model("minimal_discrete")
out <- odin2_simulate(
  mdl,
  times = 0:10,
  pars = list(inc = 1),
  initial = list(S = 0)
)

out
```

Interpretation: with `inc = 1`, the state increases linearly each step.

## 3) Susceptible-balance recurrence model

The `balance_discrete` model uses:

- `S[t+1] = max(0, S[t] + births[t] - vaccinated[t] - cases[t] / rho)`

```{r}
mdl_bal <- odin2_build_model("balance_discrete")
out_bal <- odin2_simulate(
  mdl_bal,
  times = 0:5,
  pars = list(rho = 1, births = 0, vaccinated = 0, cases = 1),
  initial = list(S = 3)
)

out_bal
```

With these parameters, susceptibles decrease by 1 each step and floor at 0.

### Time-varying inputs

```{r}
times <- 0:4
out_tv <- odin2_simulate(
  mdl_bal,
  times = times,
  pars = list(rho = 1),
  initial = list(S = 3),
  inputs = list(
    births = c(0, 1, 0, 0, 0),
    vaccinated = 0,
    cases = c(0, 0, 1, 0, 2)
  )
)

out_tv
```

## 4) End-to-end panel to mechanistic inputs

```{r}
panel_path <- system.file("extdata", "example_panel_measles_global.csv", package = "vpdsus")
panel <- utils::read.csv(panel_path)

panel_ok <- panel |>
  dplyr::filter(!is.na(iso3), !is.na(year), !is.na(births), !is.na(coverage), !is.na(cases))

iso3 <- panel_ok$iso3[[1]]
times <- sort(unique(panel_ok$year[panel_ok$iso3 == iso3]))
inputs <- odin2_balance_inputs_from_panel(panel_ok, times = times, iso3 = iso3)

inputs
```

```{r}
out_panel <- odin2_simulate(
  mdl_bal,
  times = times,
  pars = list(rho = 1),
  initial = list(S = 0),
  inputs = inputs
)

out_panel
```

## 5) Compare with Method D

`balance_discrete` is intended to match Method D under matched assumptions.

```{r}
sus_d <- estimate_susceptible_case_balance(
  panel_ok |> dplyr::filter(.data$iso3 == .env$iso3),
  rho = 1,
  s0 = 0,
  coverage_col = "coverage",
  births_col = "births",
  cases_col = "cases",
  pop_col = "pop_total",
  age_group = "total"
)

cmp <- dplyr::left_join(
  dplyr::mutate(out_panel, year = time),
  dplyr::select(sus_d, year, susceptible_n),
  by = "year"
)

cmp
```

```{r}
ggplot2::ggplot(cmp, ggplot2::aes(x = year)) +
  ggplot2::geom_line(ggplot2::aes(y = S, colour = "odin2"), linewidth = 1) +
  ggplot2::geom_point(ggplot2::aes(y = S, colour = "odin2"), size = 2) +
  ggplot2::geom_line(ggplot2::aes(y = susceptible_n, colour = "method_d"), linewidth = 1, linetype = 2) +
  ggplot2::geom_point(ggplot2::aes(y = susceptible_n, colour = "method_d"), size = 2, shape = 1) +
  ggplot2::scale_colour_manual(values = c(odin2 = "#1b9e77", method_d = "#d95f02")) +
  ggplot2::labs(
    x = "Year",
    y = "Susceptible (count)",
    colour = NULL,
    title = paste0("Susceptible balance: odin2 vs Method D (", iso3, ")")
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

## 6) Calibrate reporting fraction rho

Calibration question:

What `rho` value would make the model match a chosen susceptible count at a
target year?

```{r}
target_year <- max(times)
target_susceptible_n <- max(1, out_panel$S[[length(out_panel$S)]] * 0.9)

rho_hat <- tryCatch(
  calibrate_rho_case_balance(
    panel_ok,
    iso3 = iso3,
    years = times,
    target_year = target_year,
    target_susceptible_n = target_susceptible_n,
    rho_interval = c(0.01, 1)
  ),
  error = function(e) {
    message("Calibration failed for this panel slice; using rho = 1 for demonstration.")
    1
  }
)

rho_hat
```

```{r}
out_cal <- odin2_simulate(
  mdl_bal,
  times = times,
  pars = list(rho = rho_hat),
  initial = list(S = 0),
  inputs = inputs
)

out_cal
```

## 7) Annual SIRV model (spec-required mechanistic baseline)

The annual SIRV model includes:

- compartments `S`, `I`, `R`, `V`,
- births into `S`,
- vaccination flow `S -> V`,
- infection `S -> I`,
- recovery `I -> R`,
- mortality from all compartments,
- reporting fraction `rho` applied to infections.

```{r}
mdl_sirv <- odin2_build_model("sirv_annual")
times <- 0:15

out_sirv <- odin2_simulate(
  mdl_sirv,
  times = times,
  pars = list(
    beta = 0.8,
    gamma = 0.4,
    mu = 0.01,
    rho = 0.25,
    births = 50000,
    vaccinated = 40000,
    n_pop = 1e7
  ),
  initial = list(
    S = 2e6,
    I = 1e3,
    R = 7.95e6,
    V = 5e4
  )
)

head(out_sirv)
```

Key outputs to inspect for susceptibility inference:

- `reported_cases`: model-implied observed burden,
- `susceptible_n`: susceptible count,
- `susceptible_prop`: susceptible proportion.

```{r}
dplyr::select(out_sirv, time, reported_cases, susceptible_n, susceptible_prop) |>
  head()
```

## Scope and next step

This pathway now includes both a transparent balance model and a spec-aligned
annual SIRV model.

The next student-facing extension is to feed country-year panel inputs directly
into `sirv_annual` and compare inferred susceptibility trajectories against
Method B/Method D under matched assumptions.
