---
title: "Outbreak modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Outbreak modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette addresses a concrete research question:

Do countries with higher estimated susceptibility have higher outbreak risk in
the following year?

All examples use bundled data and run offline.

Workflow:

1. Create an outbreak outcome and lagged modelling panel.
2. Fit baseline logistic and negative-binomial models.
3. Evaluate out-of-time performance.
4. Explore sensitivity to lag and outcome definition.

Function map:

- `make_outcome_outbreak()`: turns case counts into binary outbreak labels.
- `make_modelling_panel()`: aligns predictors at year `t` with outbreak at `t + lag`.
- `fit_outbreak_models()`: fits logistic and negative-binomial baselines.
- `evaluate_outbreak_time_split()`: one train/test split by year.
- `evaluate_models_rolling_origin()`: repeated rolling-origin validation.

## 1) Build the modelling panel

```{r}
library(vpdsus)

panel <- vpdsus_example_panel()
suscept <- estimate_susceptible_static(panel, coverage_col = "coverage", pop_col = "pop_0_4")

mod_pc <- make_modelling_panel(panel, suscept, outcome = "outbreak_pc", lag_years = 1)
head(mod_pc)
```

Because this example panel is global and historical, we can inspect the modelling
sample before fitting:

```{r}
dplyr::summarise(
  mod_pc,
  n_rows = dplyr::n(),
  n_countries = dplyr::n_distinct(iso3),
  year_min = min(year, na.rm = TRUE),
  year_max = max(year, na.rm = TRUE),
  outbreak_rate = mean(outcome, na.rm = TRUE)
)
```

```{r}
mod_pc |>
  dplyr::group_by(who_region) |>
  dplyr::summarise(
    n = dplyr::n(),
    outbreak_rate = mean(outcome, na.rm = TRUE),
    mean_susceptibility = mean(susceptible_prop, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n)) |>
  head(12)
```

Key columns:

- `outcome`: binary indicator in the next year (`year_next`)
- `susceptible_prop`: main predictor
- `year`, `who_region`: adjustment terms in the baseline model

Inspect how outbreak labels are constructed from cases:

```{r}
outcomes <- make_outcome_outbreak(cases = panel$cases, pop = panel$pop_total)
head(outcomes)
```

## 2) Fit baseline outbreak model

`fit_outbreak_models()` fits:

- logistic regression for binary outbreak risk,
- negative-binomial regression for outbreak counts (`cases_next`),
- with a common covariate set: susceptibility, year, and WHO region.

```{r}
fit <- fit_outbreak_models(mod_pc)
fit$model
fit$coefficients
fit$model_nb
fit$coefficients_nb
```

### Baseline comparison: coverage-only vs susceptibility-enabled

```{r}
cmp_df <- mod_pc |>
  dplyr::filter(!is.na(outcome), !is.na(coverage), !is.na(susceptible_prop))

fit_cov <- stats::glm(outcome ~ coverage + year + who_region, data = cmp_df, family = stats::binomial())
fit_sus <- stats::glm(outcome ~ susceptible_prop + year + who_region, data = cmp_df, family = stats::binomial())
fit_both <- stats::glm(outcome ~ coverage + susceptible_prop + year + who_region, data = cmp_df, family = stats::binomial())

data.frame(
  model = c("coverage_only", "susceptibility_only", "coverage_plus_susceptibility"),
  aic = c(AIC(fit_cov), AIC(fit_sus), AIC(fit_both))
)
```

Interpretation guide:

- `estimate > 0` for `susceptible_prop` means higher susceptibility is associated
  with higher outbreak odds.
- region/year terms absorb broad structural differences and temporal trend.

## 3) Time-aware evaluation

Use [evaluate_outbreak_time_split()] for:

- a train/test split by time,
- summary metrics,
- prediction and calibration plots.

```{r}
ev <- evaluate_outbreak_time_split(mod_pc)
ev$metrics
```

Metrics meaning:

- `accuracy`: classification performance at threshold 0.5
- `brier`: probability calibration/discrimination loss (lower is better)
- `log_loss`: penalises overconfident errors
- `auc`: threshold-free discrimination

Inspect test-set predictions:

```{r}
head(dplyr::select(ev$predictions, iso3, year, year_next, outcome, prob_outbreak, pred_outbreak))
```

Diagnostic plots:

```{r}
ev$plots$time
ev$plots$calibration
```

## 4) Sensitivity checks

### A) Outcome definition: per-capita vs absolute thresholds

```{r}
mod_abs <- make_modelling_panel(panel, suscept, outcome = "outbreak_abs", lag_years = 1)
ev_abs <- evaluate_outbreak_time_split(mod_abs)

dplyr::bind_rows(
  dplyr::mutate(ev$metrics, outcome_def = "outbreak_pc"),
  dplyr::mutate(ev_abs$metrics, outcome_def = "outbreak_abs")
) |>
  dplyr::select(outcome_def, train_end, n_train, n_test, accuracy, brier, log_loss, auc)
```

### B) Lag sensitivity

```{r}
lags <- c(1, 2)
lag_metrics <- purrr::map_dfr(lags, function(l) {
  mod_l <- make_modelling_panel(panel, suscept, outcome = "outbreak_pc", lag_years = l)
  m <- tryCatch(
    evaluate_outbreak_time_split(mod_l)$metrics,
    error = function(e) {
      tibble::tibble(
        train_end = NA_real_,
        n_train = nrow(mod_l),
        n_test = NA_integer_,
        prevalence = NA_real_,
        accuracy = NA_real_,
        brier = NA_real_,
        log_loss = NA_real_,
        auc = NA_real_
      )
    }
  )
  dplyr::mutate(m, lag_years = l)
})

dplyr::select(lag_metrics, lag_years, accuracy, brier, log_loss, auc, n_test)
```

### C) Rolling-origin evaluation (recommended default for research claims)

```{r}
ev_roll <- evaluate_models_rolling_origin(mod_pc, min_train_years = 1)
ev_roll
```

### D) Regional heterogeneity snapshot

```{r}
region_perf <- ev$predictions |>
  dplyr::group_by(who_region) |>
  dplyr::summarise(
    n = dplyr::n(),
    accuracy = mean(pred_outbreak == outcome),
    prevalence = mean(outcome),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n))

region_perf
```

## 5) Conflict exposure as an additional predictor

To test whether country-year conflict exposure helps explain outbreak risk and
changes in cases/coverage, we can merge conflict indicators onto the same panel.

```{r}
conf <- get_conflict(
  years = sort(unique(panel$year)),
  countries = unique(panel$iso3),
  source = "ucdp_fixture"
)

panel_conf <- merge_conflict_with_panel(panel, conf)

conf_mod <- make_conflict_modelling_panel(
  panel_conf,
  lag_years = 1,
  outcome = "outbreak_pc",
  require_conflict_available = TRUE
)

dplyr::summarise(
  conf_mod,
  n_rows = dplyr::n(),
  conflict_share = mean(conflict_any_event, na.rm = TRUE),
  mean_conflict_fatalities_per_100k = mean(conflict_fatalities_per_100k, na.rm = TRUE)
)
```

```{r}
fit_conf <- fit_conflict_effect_models(conf_mod)

fit_conf$coefficients |>
  dplyr::filter(grepl("^conflict_", term)) |>
  dplyr::arrange(model, term)
```

## Take-home interpretation

- This module already tests an interesting and useful public-health hypothesis:
  whether susceptibility helps predict next-year outbreak risk.
- The conflict extension lets you test whether conflict occurrence (discrete)
  and intensity (continuous) add predictive information for outbreak risk and
  changes in incidence/coverage.
- The current baseline model is intentionally simple and transparent.
- Next upgrades for a research project are richer model classes, rolling-origin
  validation, and explicit baseline comparators.
