---
title: "Susceptibility estimates and WHO-style ranking plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Susceptibility estimates and WHO-style ranking plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette answers a practical first-project question:

Can we identify countries that might be at higher measles outbreak risk by
combining recent vaccination coverage with simple susceptibility estimates?

The workflow is:

1. Load and inspect a country-year panel.
2. Estimate susceptibility (Method A, Method B, and Method D).
3. Rank countries using recent coverage and current susceptibility.
4. Interpret how method choice changes the ranking.

How the key functions fit together:

- `vpdsus_example_panel()`: provides a reproducible teaching dataset.
- `panel_validate()`: checks that required columns/ranges are sane.
- `estimate_susceptible_*()`: computes susceptibility under different assumptions.
- `risk_rank()`: combines recent coverage + susceptibility into a prioritisation table.
- `plot_coverage_rank()` / `plot_susceptible_rank()`: communicate results quickly.

## 1) Load and inspect panel data

```{r}
library(vpdsus)

panel <- vpdsus_example_panel()
head(panel)
```

Before modelling, validate basic schema and ranges:

```{r}
v <- panel_validate(panel)
v$ok
v$diagnostics
v$issues
```

Check key variables used below:

```{r}
dplyr::summarise(
  panel,
  n_rows = dplyr::n(),
  n_countries = dplyr::n_distinct(iso3),
  year_min = min(year),
  year_max = max(year),
  coverage_min = min(coverage, na.rm = TRUE),
  coverage_max = max(coverage, na.rm = TRUE),
  births_min = min(births, na.rm = TRUE),
  births_max = max(births, na.rm = TRUE),
  cases_min = min(cases, na.rm = TRUE),
  cases_max = max(cases, na.rm = TRUE)
)
```

## 2) Susceptibility methods and assumptions

### Method A: static coverage-based susceptibility

Function: [estimate_susceptible_static()]

Core idea:

- susceptible proportion is approximately `1 - (coverage * VE)`.
- susceptible count is this proportion multiplied by a population denominator.

In this vignette we use:

- `coverage` as the immunisation proportion,
- `pop_0_4` as the denominator for a simple child-age signal,
- `VE = 1` (default) for a minimal didactic example.

```{r}
sA <- estimate_susceptible_static(
  panel,
  coverage_col = "coverage",
  pop_col = "pop_0_4"
)

head(dplyr::select(sA, iso3, year, susceptible_prop, susceptible_n, method))
```

### Method B: cohort reconstruction from births and coverage

Function: [estimate_susceptible_cohort()]

Core idea:

- reconstruct the recent birth cohorts that make up an age window (default 0-4),
- for each cohort, count births not effectively immunised,
- aggregate across the window and divide by the age-group population.

Equation for year `y` with `years_back = 4`:

- `S_n(y) = sum_{k=0..4} births(y-k) * (1 - coverage(y-k))`
- `S_prop(y) = S_n(y) / pop_0_4(y)`

So Method B explicitly answers:
"How many children currently aged 0-4 likely remained susceptible, given their
birth cohort sizes and the coverage observed when each cohort was vaccinated?"

### Method B walkthrough (single country, step by step)

To make the logic concrete, we take one country and one year and rebuild the
cohort sum by hand.

```{r}
demo_candidates <- panel |>
  dplyr::filter(!is.na(coverage), !is.na(births), !is.na(pop_0_4)) |>
  dplyr::count(iso3, name = "n_years") |>
  dplyr::filter(n_years >= 5) |>
  dplyr::arrange(dplyr::desc(n_years))

iso_demo <- demo_candidates$iso3[[1]]
year_demo <- panel |>
  dplyr::filter(iso3 == iso_demo, !is.na(coverage), !is.na(births), !is.na(pop_0_4)) |>
  dplyr::summarise(year_demo = max(year)) |>
  dplyr::pull(year_demo)

cohort_tbl <- panel |>
  dplyr::filter(iso3 == iso_demo, year %in% (year_demo - 4):year_demo) |>
  dplyr::arrange(year) |>
  dplyr::transmute(
    iso3,
    year,
    births,
    coverage,
    susceptible_from_cohort = births * (1 - coverage)
  )

cohort_tbl
```

Manual Method B numerator for this example year:

```{r}
manual_s_n <- sum(cohort_tbl$susceptible_from_cohort, na.rm = TRUE)
manual_s_n
```

Compare manual reconstruction with package output:

```{r}
sB <- estimate_susceptible_cohort(
  panel,
  coverage_col = "coverage",
  births_col = "births",
  pop_col = "pop_0_4"
)

head(dplyr::select(sB, iso3, year, susceptible_prop, susceptible_n, method))
```

```{r}
sB_demo <- sB |>
  dplyr::filter(iso3 == iso_demo, year == year_demo) |>
  dplyr::select(iso3, year, susceptible_n, susceptible_prop)

manual_check <- panel |>
  dplyr::filter(iso3 == iso_demo, year == year_demo) |>
  dplyr::transmute(
    iso3,
    year,
    pop_0_4,
    manual_susceptible_n = manual_s_n,
    manual_susceptible_prop = manual_s_n / pop_0_4
  ) |>
  dplyr::left_join(sB_demo, by = c("iso3", "year"))

manual_check
```

On a large global panel this method is often much more informative than Method A,
because it uses cohort size and birth-history dynamics directly rather than only
current coverage.

### Method D: case-balance recurrence with reporting fraction rho

Function: [estimate_susceptible_case_balance()]

Core recurrence (discrete-time):

- `S[t+1] = max(0, S[t] + births[t] - vaccinated[t] - cases[t] / rho)`

Interpretation:

- lower `rho` means more under-reporting and therefore larger implied depletion
  from infection for the same observed cases.

```{r}
sD <- estimate_susceptible_case_balance(
  panel,
  rho = c(0.2, 0.5, 1),
  coverage_col = "coverage",
  births_col = "births",
  cases_col = "cases",
  pop_col = "pop_total",
  age_group = "total"
)

head(dplyr::select(sD, iso3, year, rho, susceptible_prop, susceptible_n, method))
```

Choosing a method for a first project:

- Start with Method A to understand the workflow quickly.
- Move to Method B when births and cohort structure matter.
- Use Method D when reported cases are considered informative about depletion.

## 3) Rank countries and produce WHO-style outputs

The ranking helper [risk_rank()] combines:

- recent mean coverage over a window (`window_years`), and
- susceptibility at a target end year (`year_end`).

```{r}
rank_A <- risk_rank(panel, sA, window_years = 3, year_end = 2020)
rank_B <- risk_rank(panel, sB, window_years = 3, year_end = 2020)

head(dplyr::select(rank_A, iso3, coverage_mean, susceptible_prop, risk_category))
```

Plots:

```{r}
plot_coverage_rank(rank_A, top_n = 5)
plot_susceptible_rank(rank_A, top_n = 5)
```

## 4) How method choice affects interpretation

Compare trajectories for one country:

```{r}
compare_methods(
  panel = panel,
  estimates = list(
    static = sA,
    cohort = sB
  ),
  iso3 = panel$iso3[[1]]
)
```

Simple side-by-side ranking comparison:

```{r}
dplyr::left_join(
  dplyr::mutate(
    dplyr::select(rank_A, iso3, risk_A = risk_category, susc_A = susceptible_prop),
    year_end = attr(rank_A, "year_end")
  ),
  dplyr::mutate(
    dplyr::select(rank_B, iso3, risk_B = risk_category, susc_B = susceptible_prop),
    year_end = attr(rank_B, "year_end")
  ),
  by = c("iso3", "year_end")
)
```

## Take-home interpretation

- Method A is the easiest baseline and useful for quick screening.
- Method B adds births/cohort structure and is usually more epidemiologically
  interpretable for child susceptibility.
- Method D introduces case-driven depletion and a reporting parameter (`rho`),
  bridging towards mechanistic inference.
